---
title: "STAT 605 Progress Report"
author: "Bing Xu"
date: '`r format(Sys.time(), "%m/%d/%Y")`'
output: pdf_document
---

```{r setup,include=FALSE}
require(tidyverse)
require(stringr)
require(forcats)
require(ggplot2)
require(choroplethr)
require(choroplethrMaps)
require(dplyr)
library(tidyverse)

read.csv("large_sample_races.csv", header = TRUE, stringsAsFactors = FALSE) %>%
  tbl_df() ->
  A

#Read in County FIPS Key
read.csv("fips_codes_website.csv", header = TRUE, stringsAsFactors = FALSE) %>%
  tbl_df() ->
  B

#convert to tibble so dplyr can work its magic
A_df = tbl_df(A)

#format FIPs in source file!
head(A_df)
A_df_char = mutate(A_df, SFIP_char = as.character(STATEFIP), CFIP_char = as.character(COUNTYFIPS))
A_df_char$SFIP_char = str_pad(A_df_char$SFIP_char, 2, pad = "0")
A_df_char$CFIP_char = str_pad(A_df_char$CFIP_char, 3, pad = "0")
head(A_df_char)
A_df_char = mutate(A_df_char, FIPS = paste(SFIP_char, CFIP_char, sep = ""))

head(B)
B2 = filter (B, Entity.Description == "County")
head(B2)
B2$State.FIPS.Code = as.character(B2$State.FIPS.Code)
B2$County.FIPS.Code = as.character(B2$County.FIPS.Code)
B2$State.FIPS.Code = str_pad(B2$State.FIPS.Code, 2, pad = "0")
B2$County.FIPS.Code = str_pad(B2$County.FIPS.Code, 3, pad = "0")
B2 = mutate(B2,FIPS = paste(State.FIPS.Code,County.FIPS.Code, sep = ""))
head(B2)
filter(B2, FIPS == "15003")


#RACE CODES
# 1 = white
# 2 = black
# 3 = American Indian/Alaskan Native
# 4 = Chinese
# 5 = Japanese
# 6 = Other Asian / Pacific Islander
# 7 = Othe Race, nec
# 8 = Two major races
# 9 = Three or more races

#HAVE TO ADD COLUMNS FOR EACH RACE, WHICH WILL MAKE SUMMARIZING EASIER
A2=A_df_char %>% mutate(W = ifelse(RACE==1,1,0))
A2 = A2 %>% mutate(B = ifelse(RACE==2,1,0))
A2 = A2 %>% mutate(AI = ifelse(RACE==3,1,0))
A2 = A2 %>% mutate(C = ifelse(RACE==4,1,0))
A2 = A2 %>% mutate(J = ifelse(RACE==5,1,0))
A2 = A2 %>% mutate(PI = ifelse(RACE==6,1,0))
A2 = A2 %>% mutate(O = ifelse(RACE==7 | RACE == 8 | RACE == 9 ,1,0))


#summarise race rsults and create diversity index
by_county = A2 %>% group_by(FIPS) %>% dplyr::summarise( tot = n(), W = sum(W)/tot, B = sum(B)/tot, AI = sum (AI)/tot, C = sum(C)/tot, J = sum(J)/tot, PI = sum(PI)/tot, O = sum(O)/tot)
by_county = mutate(by_county,TOT = W+B+AI+C+J+PI+O)
by_county = by_county %>% mutate(DiversityIndex = 1 - W^2-B^2-AI^2-C^2 -J^2 - PI^2 - O^2) #calculate diversity indec
by_county = arrange(by_county, desc(DiversityIndex)) #arrange in descenting order
by_county

#set up data to be mapped by choroplethR
mapping = select(by_county, FIPS, DiversityIndex) 
mapping = dplyr::rename(mapping, region=FIPS, value=DiversityIndex) 
mapping$region = as.numeric(mapping$region)

#map data
county_choropleth(mapping, title = "Diversity Index in Counties Across the US", state_zoom=c("california"))
#can also add optiions "state_zoom = c("california") or other lowercase state names

head(by_county)

C = left_join(by_county, B2, by = "FIPS")
C

C2 = select(C, FIPS, DiversityIndex, State.Abbreviation, GU.Name)
C2

##Importing Education dataset
read.csv("Education.csv", header = TRUE, stringsAsFactors = FALSE) %>%
  tbl_df() ->
  edu

##Merge two datasets
by_county$FIPS<-as.numeric(by_county$FIPS)
edu_comb<-left_join(by_county, edu, by=c("FIPS"="FIPS.Code"))
```

We also find a data set that include the percentage of attaining certain education level for each county. So we are interested in finding whether there are relationships bewtween the diversity we devised and percentages of completing degrees. There are four percentages, which are percent of adults with less than a high school diploma, percent of adults with a high school diploma only, percent of adults completing some college or associate's degree, percent of adults with a bachelor's degree or higher.  

```{r,echo=FALSE, warning=FALSE,fig.height=4,fig.width=6}
##Plot
edu_comb %>% 
  ggplot(aes(x=DiversityIndex,
             y=Percent.of.adults.with.less.than.a.high.school.diploma..2010.2014))+
  geom_point()+geom_smooth()+
  ylab("Percentage")+
  xlab("Diversity Index")+
  ggtitle("Diversity vs No High School Diploma Percentage")

edu_comb %>% 
  ggplot(aes(x=DiversityIndex,
             y=Percent.of.adults.with.a.high.school.diploma.only..2010.2014))+
  geom_point()+geom_smooth()+
  ylab("Percentage")+
  xlab("Diversity Index")+
  ggtitle("Diversity vs High School Diploma Only Percentage")


edu_comb %>% 
  ggplot(aes(x=DiversityIndex,
             y=Percent.of.adults.completing.some.college.or.associate.s.degree..2010.2014))+
  geom_point()+geom_smooth()+
  ylab("Percentage")+
  xlab("Diversity Index")+
  ggtitle("Diversity vs Completing College or Associate Percentage")

edu_comb %>% 
  ggplot(aes(x=DiversityIndex,
             y=Percent.of.adults.with.a.bachelor.s.degree.or.higher..2010.2014))+
  geom_point()+geom_smooth()+
  ylab("Percentage")+
  xlab("Diversity Index")+
  ggtitle("Diversity vs Bachelor Degree or Higher Percentage")

##Linear Regression Analysis
l1<-lm(Percent.of.adults.with.less.than.a.high.school.diploma..2010.2014~DiversityIndex, data = edu_comb)
##summary(l1)
l2<-lm(Percent.of.adults.with.a.high.school.diploma.only..2010.2014~DiversityIndex, data = edu_comb)
##summary(l2)
l3<-lm(Percent.of.adults.completing.some.college.or.associate.s.degree..2010.2014~DiversityIndex, data = edu_comb)
##summary(l3)
l4<-lm(Percent.of.adults.with.a.bachelor.s.degree.or.higher..2010.2014~DiversityIndex, data = edu_comb)
##summary(l4)
```

From graphs above we can see that all rates seems not to have strong correlation with the diversity index. We also conducted the linear regression analysis and all regressions are having a relatively poor $R^{2}$, which the maximum is 13.27% (The regression between diversity index and the percent of completing a high school diploma only).  

We can find that there is a decreasing trend between the diversity index and the percentage of adults completing a high school degree only. We can find that counties whose high school diploma only percentage is high are having a relatively lower diversity index, which means that the race are more concentrated in these counties. 
As for the relationship between the percentage of less than a high school diploma and the diversity index, the trend is a bit oppoiste. Highly diversified counties are a bit likely to have a higher percentage. There are some outliers and these outliers implies that counties with either low or high diversity indexes could have a high percentage.

While for the last two graphs, the relationships between variables are quite weak and there are many outliers in degree percentages.


