---
title: "Effects of County-Level Diversity on Educational, Economic,and Health-Related Outcomes"
author: Kenny Groszman, Jack Petersen, Bing Xu
output: pdf_document
fontsize: 10pt
geometry: margin=1in
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Introduction:

It is often touted by the government, businesses, and members of our community that diversity brings with it a variety of benefits [1]. Indeed, it has been shown that at the organizatin-level, diversity does provide micro-scale benefits such as new perspectives, increased productivity, expanded worldviews for the members of the organization, and more [2]. This effect leads us to ask the following question: *does diversity bring with it macro-scale benefits? More specificially, is being in a more diverse county correlated with statistical indicators of success in a variety of different categories?*

We will consider those three categories of possible downstream effects of diversity: educational, economic, and health-related outcomes. It is possible that increased diversity will actually have negative effects on a broader scale, as certain minority communities may experience alienation. However, it is also possible that different racial groups will live harmoniously, learn from each other, and contribute to an overall more successful community.

It should also be noted that right off the bat, these categories of success that we are looking at are affected by a plethora of factors that do not include racial diversity (geographic location, history, politics, major industries and employers, etc.). Thus, we expect that correlations will be weak. However, we also expect to be able to glean some insights about the effects of broad diversity in America.

## Data

The approach to data collection for this project was to take several different county-level data sets and to join them using the unique  Federal Information Processing Standard (FIPS) code for each county. Education, Health, and Economics data were taken from county-level databases published by several different government agencies. However, county diversity and race data were constructed from a household-level database called the Integrated Public Use Microdata Series (IPUMS), which was then grouped by county for further analysis.

The variables we were able to pull from these databases and use for analysis, as well as their sources, are below:

  
Variable | Description |Source
------------- | --------------------------------------------------------- | -----
Geographic Location | County location of record, given by state and county FIPS (Federal Info. Processing Standard ) code | [3]
Race | Numerical code corresponding to race of individual: white, black,  American Indian, Chinese, Japanese, Pacific Islander, other, 2+ races | [3]
Educational Attainment | Code corresponding to highest level of education achieved: no high school,  high school diploma,  Associate's degree, Bachelor's degree | [4]
Poverty Status | Code indicating whether individual falls below the poverty line | [5]
Income | Integer indicating the household's yearly income | [5]
Life Expectancy | Database of life expectancy for differentcounties in the US | [6]
Obesity | Rates of obesity in different US counties | [6]


## Creating a Diversity Index
Upon summarizing race data by county, race information is given as the proportion of the county population that falls into 8 different groups:

* White
* Black
* American Indian/Alaskan Native
* Chinese
* Japanese
* Other Asian / Pacific Islander
* Othe Race
* 2+ races

We needed a measure of Diversity that follows the following conditions:

1. The county with the highest possible Diversity Index Score is one with the population spread evenly across the 8 groups (i.e. 12.5% in each group)
2. A county with the lowest possible Diversity Index Score is one with the population entirely in one of the groups (i.e. 100% White, 0% everything else)

A formula that meets this criteria is for county *i* is as follows, for races $R \in \{White, Black,...,  2+ Races\}$ :

$$ DiversityIndex_i = 1 -  \sum_{j \in R} x_{ij}^{2} $$

where $x_{ij}$ is the proportion of race *j* in county *i* [7].

### Results and Interpretation

```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.height=4}
library(tidyverse)
library(stringr)
library(forcats)
library(ggplot2)
library(choroplethr)
library(choroplethrMaps)
library(dplyr)

setwd("C:/Users/kenny/OneDrive/Documents/Fall 2016/STAT 405- R for Data Science/STAT_405_Diversity_Project/Final Paper/Final Data/")

dat<-read.csv("CC-EST2015-ALLDATA.csv",header = TRUE)
#head(dat)

##FOR ALL AGE GROUPS AND FOR YEAR 2015
dat2015<-dat %>% 
  filter(AGEGRP==0 & YEAR==8)

##DO NOT COUNT HISPANIC
##SIMPLY USING TWO 6 TYPES OF RACES AND THIS COULD BE REVISED 
##H=H_MALE+H_FEMALE,
dat2015<-dat2015 %>% 
  transmute(STATE=STATE,
            COUNTY=COUNTY,
            TOT=TOT_POP,
            WA=WA_MALE+WA_FEMALE,
            BA=BA_MALE+BA_FEMALE,
            IA=IA_MALE+IA_FEMALE,
            AA=AA_MALE+AA_FEMALE,
            NHA=NA_MALE+NA_FEMALE,
            MORE=TOT-WA-BA-IA-AA-NHA,
            WAP=WA/TOT,
            BAP=BA/TOT,
            IAP=IA/TOT,
            AAP=AA/TOT,
            NHAP=NHA/TOT,
            MOREP=MORE/TOT,
            INDEX=1-WAP^2-BAP^2-IAP^2-AAP^2-NHAP^2-MOREP^2)

dat2015<-dat2015 %>% 
  mutate(STATE_char = as.character(STATE), COUNTY_char = as.character(COUNTY))
dat2015$STATE_char = str_pad(dat2015$STATE_char, 2, pad = "0")
dat2015$COUNTY_char = str_pad(dat2015$COUNTY_char, 3, pad = "0")
#head(dat2015)
dat2015<-dat2015 %>% 
  mutate(FIPS = paste(STATE_char, COUNTY_char, sep = ""))
#head(dat2015)

mapping <- select(dat2015, FIPS, INDEX) 
mapping <- dplyr::rename(mapping, region=FIPS, value=INDEX) 
mapping$region <- as.numeric(mapping$region)

county_choropleth(mapping, title = "Diversity Index in Counties Across the US")


```

We see from this map that diverse counties in the US are concentrated near the conty's southern, western, and eastern borders. Additionally, Alaska is quite diverse. To the contrary, the midwest and central USA are among the lowest diversity counties. These results are important to keep in mind as we analyze downstream effects, as particulars about these regions may confound the analysis. For example, if we were to look at hospital quality based on diversity, our results might be skewed since the best hospitals in the country are in California and New York, for reasons that are probably independent of racial diversity. 

Another interesting note is that there are far more counties with quite low levels of diversity (0 - 0.2) than there are counties with higher levels (0.2 - 0.6). There does seem to be a spike at arounf 0.5, and this might be doe to large coastal counties. Very few counties have a diversity index greater than 0.6.

```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.height=3}
ggplot(dat2015, aes(x=INDEX)) + geom_histogram(stat="bin", aes(fill= ..count..), bins =50) + 
  xlab("Diversity Index") + ylab("Number of Counties with Index") + ggtitle("Diversity Index Distribution in US Counties")

```

## Effects of Diversity on Educational Outcomes

```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.height=4,eval=FALSE}


```



## Effects of Diversity on Economic Outcomes

In this section, we are going to look at whether there is a relationship between the diversity and economic outcomes within counties. We are using variables like poverty rate, unemployment rate, median income and the income gap as indicators of counties' economic outcomes.

### *1. US Census Bureau Poverty Estimates*

US Census Bureau Model-based Small Area Income & Poverty Estimates provides the estimates of the total population and the population of falling below the poverty line in each county in 2014 and we get the poverty rate estimates from this dataset. And we categorize counties into four types by the diversity index. The four types are:

* Level 1: The diversity index is from 0 to 0.2.
\newline
* Level 2: The diversity index is from 0.2 to 0.4.
\newline
* Level 3: The diversity index is from 0.4 to 0.6.
\newline
* Level 4: The diversity index is from 0.6 to 0.8.

Then we draw the following graph and try to find whether there are differences of poverty rates in different diversity index ranges.


```{r,echo=FALSE,message=FALSE,warning=FALSE,fig.height=4}

setwd("C:/Users/kenny/OneDrive/Documents/Fall 2016/STAT 405- R for Data Science/STAT_405_Diversity_Project/Final Paper/Final Data/")
dat<-read.csv("CC-EST2015-ALLDATA.csv",header = TRUE)

dat2014<-dat %>% 
  filter(AGEGRP==0 & YEAR==7)
dat2014<-dat2014 %>% 
  transmute(STATE=STATE,
            COUNTY=COUNTY,
            TOT=TOT_POP,
            WA=WA_MALE+WA_FEMALE,
            BA=BA_MALE+BA_FEMALE,
            IA=IA_MALE+IA_FEMALE,
            AA=AA_MALE+AA_FEMALE,
            NHA=NA_MALE+NA_FEMALE,
            TOM=TOM_MALE+TOM_FEMALE,
            WAP=WA/TOT,
            BAP=BA/TOT,
            IAP=IA/TOT,
            AAP=AA/TOT,
            NHAP=NHA/TOT,
            TOMP=TOM/TOT,
            INDEX=1-WAP^2-BAP^2-IAP^2-AAP^2-NHAP^2-TOMP^2)

dat2014<-dat2014 %>% 
  mutate(STATE_char = as.character(STATE), COUNTY_char = as.character(COUNTY))
dat2014$STATE_char = str_pad(dat2014$STATE_char, 2, pad = "0")
dat2014$COUNTY_char = str_pad(dat2014$COUNTY_char, 3, pad = "0")
dat2014<-dat2014 %>% 
  mutate(FIPS = paste(STATE_char, COUNTY_char, sep = ""))
dat2014$FIPS<-as.numeric(dat2014$FIPS)

poverty<-read.csv("PovertyEstimates.csv", header = TRUE, stringsAsFactors = FALSE)

dat2014b<-dat2014 %>% 
  mutate(group=0)

  for (i in 1:nrow(dat2014b)) {
    if(dat2014b$INDEX[i]>0 & dat2014b$INDEX[i]<0.2)
      dat2014b$group[i]="Level 1"
    else if (dat2014b$INDEX[i]>=0.2 & dat2014b$INDEX[i]<0.4)
      dat2014b$group[i]="Level 2"
    else if (dat2014b$INDEX[i]>=0.4 & dat2014b$INDEX[i]<0.6)
      dat2014b$group[i]="Level 3"
    else
      dat2014b$group[i]="Level 4"
  }

poverty_comb2014b<-left_join(dat2014b, poverty, by=c("FIPS"="FIPStxt"))
poverty_comb2014b %>% 
  ggplot(aes(x=group,
             y=PCTPOVALL_2014))+
  geom_boxplot()+
  ylab("Poverty Rate")+
  xlab("Diversity Index")+
  ggtitle("Poverty Rates for Counties in Different Diversity Index Ranges in 2014")
```

#### *Commentary and Interpretation:*

We can observe that when a county's diversity index is extremely low or high, its poverty rate tends to be relatively low. For counties that have an intermediate diversity index, the poverty rates are increasing. The poverty rates of Level 3 are significantly higher than other groups'. 

One possible explanation is that in counties with intermediate diversity levels, some minority groups are possibly marginalized and these groups are more easily to be in poverty status which raise the poverty rate. While in slightly or highly diversified counties, this situation will be unlikely to occur.

### *2. US Census Bureau Median Household Income*

From US Census Bureau Model-based Small Area Income & Poverty Estimates, we can also get the annual median household income data for each county in 2014. 

We create the similar boxplot using the median household income and the diversity index level. 

```{r,warning=FALSE,echo=FALSE,fig.height=4}
setwd("C:/Users/kenny/OneDrive/Documents/Fall 2016/STAT 405- R for Data Science/STAT_405_Diversity_Project/Final Paper/Final Data/")

eco<-read.csv("Unemployment_and_Income1.csv",header = TRUE, stringsAsFactors = FALSE)
eco$Median_Household_Income_2014<-as.numeric(eco$Median_Household_Income_2014)
eco_comb2014b<-left_join(dat2014b, eco, by=c("FIPS"="FIPS_Code"))
eco_comb2014b %>% 
  ggplot(aes(x=group,
             y=Median_Household_Income_2014))+
  geom_boxplot()+
  ylab("Median Income")+
  xlab("Diversity Index")+
  ggtitle("Median Income for Different Diversity Levels in 2014")
```

#### *Commentary and Interpretation:*

We can see that the median household income for Level 3 is lower than other groups', which is consistent with results in the poverty rate part. So the low median household income for Level 3 may be due to the marginalization of some minority groups. 

We can also notice that Level 4's median household income is the higher than other three groups'. One possible reason is that highly diversified counties are more international and more international regions are more likely to have higher income levels.

### *3. IPUMS Educational Attainment*

We also use the IPUMS dataset to observe the relationship between the income gap and the diversity index. We use the interquartile range of household incomes for a given county as the income gap.

```{r,warning=FALSE,echo=FALSE,fig.height=4}
setwd("C:/Users/kenny/OneDrive/Documents/Fall 2016/STAT 405- R for Data Science/STAT_405_Diversity_Project/Final Paper/Final Data/")
  
read.csv("economics_ipums_extract.csv", header = TRUE, stringsAsFactors = FALSE) %>%
  tbl_df() ->
  A

#Read in County FIPS Key
read.csv("fips_codes_website.csv", header = TRUE, stringsAsFactors = FALSE) %>%
  tbl_df() ->
  B

#convert to tibble so dplyr can work its magic
A_df = tbl_df(A)

#format FIPs in source file!
A_df_char = mutate(A_df, SFIP_char = as.character(STATEFIP), CFIP_char = as.character(COUNTYFIPS))
A_df_char$SFIP_char = str_pad(A_df_char$SFIP_char, 2, pad = "0")
A_df_char$CFIP_char = str_pad(A_df_char$CFIP_char, 3, pad = "0")
A_df_char = mutate(A_df_char, FIPS = paste(SFIP_char, CFIP_char, sep = ""))

B2 = filter (B, Entity.Description == "County")
B2$State.FIPS.Code = as.character(B2$State.FIPS.Code)
B2$County.FIPS.Code = as.character(B2$County.FIPS.Code)
B2$State.FIPS.Code = str_pad(B2$State.FIPS.Code, 2, pad = "0")
B2$County.FIPS.Code = str_pad(B2$County.FIPS.Code, 3, pad = "0")
B2 = mutate(B2,FIPS = paste(State.FIPS.Code,County.FIPS.Code, sep = ""))

#HAVE TO ADD COLUMNS FOR EACH RACE, WHICH WILL MAKE SUMMARIZING EASIER
A2=A_df_char %>% mutate(W = ifelse(RACE==1,1,0))
A2 = A2 %>% mutate(B = ifelse(RACE==2,1,0))
A2 = A2 %>% mutate(AI = ifelse(RACE==3,1,0))
A2 = A2 %>% mutate(C = ifelse(RACE==4,1,0))
A2 = A2 %>% mutate(J = ifelse(RACE==5,1,0))
A2 = A2 %>% mutate(PI = ifelse(RACE==6,1,0))
A2 = A2 %>% mutate(O = ifelse(RACE==7 | RACE == 8 | RACE == 9 ,1,0))


#summarise race rsults and create diversity index
by_county = A2 %>% group_by(FIPS) %>% dplyr::summarise( tot = n(), W = sum(W)/tot, B = sum(B)/tot, AI = sum (AI)/tot, C = sum(C)/tot, J = sum(J)/tot, PI = sum(PI)/tot, O = sum(O)/tot)
by_county = mutate(by_county,TOT = W+B+AI+C+J+PI+O)
by_county = by_county %>% mutate(DiversityIndex = 1 - W^2-B^2-AI^2-C^2 -J^2 - PI^2 - O^2) #calculate diversity indec
by_county = arrange(by_county, desc(DiversityIndex)) #arrange in descenting order
mapping = select(by_county, FIPS, DiversityIndex) 


A_df_char %>% dplyr::group_by(FIPS) %>% 
  dplyr::summarise(q1 = quantile(POVERTY, probs = .25),
                   q3= quantile(POVERTY, probs = .75),
                   IQR = q3-q1) -> C

D = dplyr::inner_join(C, mapping, by = c("FIPS"))

fit = lm(formula =D$IQR ~ D$DiversityIndex)

#Create point graph

D %>% 
ggplot(aes(x=DiversityIndex, y=IQR)) +geom_point() +  
labs(title="Income Gap Varies with Diversity", x="Diversity Index", y="IQR") +
geom_smooth(method=lm, se=TRUE) + 
  annotate("text",x=0.5,y=460,label="y=290.994+56.665x")
```

We observe a positive relationship between the income gap and the diversity index from the scatterplot. The correlation is not very strong as the income gap could be affected by many factors. A county with a higher diversity index tends to have a more obvious income gap. This trend may be explained by the racial divide between highest and lowest-paying jobs, which increase the income gap. While for less diverse counties, this phenomenon is not that severe. 


## Effects of Diversity On Health-Related Outcomes

```{r,echo=FALSE,warning=FALSE,message=FALSE,fig.height=4,eval=FALSE}


```



## Future Plans and Conclusions

### Future Plans
As always, there is more analysis that could be done that was outside the scope of this project. For example, it would be useful to control for racial differences in outcomes (for example, differences in obesity rates in white vs. black Americans). Thus, it would be helpful to look at the different performance of the different races in counties with different diversity levels. It might also be useful to look at this data in the reverse order. For example, we could ask the question "do people with higher incomes tend to be in more or less diverse counties?". This might give us some additional insights about causation that our current analysis is lacking. With more time and more data, both of these research directions would provide extremely interesting results.

### Conclusion

In all, county-level diversity is indeed correlated with certain measures of educational, economic, and health-related outcomes. However, the analysis is quite limited since many other external factors also play into these outcomes (size of county, location, major industries, urbanization status, etc.). In general, we see a certain tradeoff between diversity and beneficial outcomes. For example, in the economic section, we saw that intermediate levels of diversity were associated with higher poverty rates and lower income, and we posit that this might be caused by certain racial groups being marginalized in these counties. This is bolstered by the observation that at the highest diversity levels, the poverty rates go back down. This might be because if a certain minority gorup is present in high enough proportions, this marginalization does not occur. Similarly, obesity is also highest in intermediate-diversity counties. 

This leads us to hypothesize that living in an extremely undiverse or an extremely diverse county results in better outcomes, and this might just be because in these counties, one has a community of people bery similar to themselves on which to rely on and feel comfortable with. To the contrary, in counties with many different groups present in small numbers, outcomes are generally worse.

With that, it looks like the groups who tout the categorical benefits of diversity are only partially correct.


